---
title: "Intermediate R for Social Science"
author: "Miruna Barnoschi"
date: "October 30^th^, 2019"
output:
  pdf_document:
    toc: yes
  html_document:
    highlight: tango
    toc: yes
    toc_float: yes
---

# Overview

The goal of this intermediate R workshop is to start building the skills to (1) better utilize R Studio for social science research, (2) visualize data using the `ggplot2` package in R, (3) transform data using the `dplyr` package in R, and (4) use data visualization and transformation skills to explore social science data in a systematic way. The latter goals are first step into `tidyverse`, a suite of packages designed for data science that are immensely helpful when doing social science work.
<br><br>

# Resources

There are several resources that would be helpful:  
First, there is \textit{R for Data Science} by Hadley Wickham and Garrett Grolemund, which is useful for learning to visualize, model, transform, tidy, and import data in a more formal way. It is free online: https://r4ds.had.co.nz.  
Second, there is "stack overflow", which can be a lifesaver for any question that comes up with respect to R (see R tags): https://stackoverflow.com. If you have a question, there is someone who has probably asked it on stack overflow and got an answer.  
Third, there is "r-bloggers", which can be helpful with respect to R explanations and tutorials: https://www.r-bloggers.com.  
Fourth, there are the always handy R cheat sheets: https://rstudio.com/resources/cheatsheets/.  
Fifth, there is of course google (a simple google search of the issue/error often does the trick).
<br><br>

# R Packages

```{r, message=FALSE}
library(tinytex) 
# package that provides LaTeX distribution for R:
# https://cran.r-project.org/web/packages/tinytex/tinytex.pdf
library(tidyverse) 
# set of packages for data science:
# https://cran.r-project.org/web/packages/tidyverse/tidyverse.pdf
library(readr) 
# package that provides a way to read data in R:
# https://cran.r-project.org/web/packages/readr/index.html
library(ggplot2)
# package that provides a way to create graphics:
# https://cran.r-project.org/web/packages/ggplot2/ggplot2.pdf
library(dplyr)
# package that provides a way to manipulate data:
# https://cran.r-project.org/web/packages/dplyr/dplyr.pdf
library(kableExtra)
# package to build complex HTML/LaTeX tables:
# https://cran.r-project.org/web/packages/kableExtra/kableExtra.pdf
library(gridExtra)
# package to arrange multiple grid-based plots on a page:
# https://cran.r-project.org/web/packages/gridExtra/gridExtra.pdf
```
<br><br>

# R Project

One best practice organizationally is to create a new R Project in an existing directory (presumably in the folder where the data is saved and the research notes/paper using that data is to be saved). Doing so saves time in terms of setting a working directory every time one works on this research project; more importantly, doing so makes replication easier.
<br><br>

# Dataset

We will look at a political science dataset that would be of interest to researchers in the subfields of comparative politics and international relations: ["UCDP Battle-related Deaths Dataset Version 19.1"](https://ucdp.uu.se/downloads/).
<br><br>

# Read Data into R

With the `readr` package, we can read in `BattleDeaths_v19_1.csv` using the appropriate function from the package, namely `read_csv()`. Datasets are usually "delimited" and `read_csv()` and `read_tsv()` are special cases of the general `read_delim()`, useful for reading comma separated values and tab separated values (i.e. the most common types of file data). There are other packages that can read data into R. With the `readxl` package, we can read in Excel fils (both .xls and .xlsx) using `read_excel()`. With the `haven` package, we can read in SPSS, Stata, and SAS files using `read_dta()`, `read_stata()`, `raed_sas()`, etc.

```{r, message=FALSE, warning=FALSE}
data <- read_csv("BattleDeaths_v19_1.csv")
```
<br><br>

# Filtering, Selecting, and Arranging Data

The war in Afghanistan has gone on from 2001 till today. Let's examine more specifically battle deaths in Afghanistan from 2001 till today to get a better picture of the conflict in that location. We would do this by filtering the data by location, "Afghanistan", and by year (2001-present) and selecting the variables associated with battle deaths (`bd_best`, `bd_low`, `bd_high`).

```{r}
data %>%
  filter(battle_location == "Afghanistan" & year >= 2001) %>%
  arrange(year) %>%
  select(conflict_id, battle_location, year, bd_best, bd_low, bd_high)
```
<br>

We can change the R output of the data (tibble) to a more formatted table.

```{r}
data %>%
  filter(battle_location == "Afghanistan", year >= 2001) %>%
  arrange(year) %>%
  select(conflict_id, battle_location, year, bd_best, bd_low, bd_high) %>%
  kable(col.names = c("Conflict ID",
                      "Battle Location",
                      "Year",
                      "Battle Deaths (best estimate)",
                      "Battle Deaths (low estimate)",
                      "Battle Deaths (high estimate)")) %>%
  kable_styling(latex_options="scale_down") # table is too big for page without this
```
<br><br>

# Summary Statistics

We can find the summary statistics (e.g. mean) of battle deaths in the subset of the dataset that focuses on the time period of the war in Afghanistan. We can also do other calculations in R (e.g. addition/sum)

```{r}
Afghanistandata <- data %>%
  filter(battle_location == "Afghanistan", year >= 2001) %>%
  arrange(year) %>%
  select(conflict_id, battle_location, year, bd_best, bd_low, bd_high)

summary(Afghanistandata$bd_best)
summary(Afghanistandata$bd_low)
summary(Afghanistandata$bd_high)

mean(Afghanistandata$bd_best)
mean(Afghanistandata$bd_low)
mean(Afghanistandata$bd_high)

sum(Afghanistandata$bd_best)
sum(Afghanistandata$bd_low)
sum(Afghanistandata$bd_high)
```
<br>

\textit{As can be seen from the table displaying the Afghanistan subset of the data corresponding to the war in Afghanistan time period, there were three distinct conflicts at that time. Should that matter for our battle deaths calculations? It is always important to know the exact data that should be analyzed prior to doing any analysis.}

*Do it yourself*: How would you examine battle deaths in the war in Iraq, which lasted from 2003 till 2011?
<br><br>

# Finding Data

Let's assume we do not want to limit ourselves to looking at data during the war in Afghanistan, but instead at data that involves Afghanistan (regardless if battles deaths occurred in Afghanistan per se). We could do so by filtering each variable with "Afghanistan". Or, more simply, we could filter all the data for any variables with the string pattern, "Afghanistan". There are numerous handy shortcuts in `dplyr` that can help us identify and find subsets of the data/specific values and information in the data.

```{r}
data %>%
  filter_all(any_vars(str_detect(., pattern = "Afghanistan"))) %>%
  arrange(year) %>%
  select(conflict_id, location_inc, side_a, side_a_2nd, side_b, side_b_2nd, year, battle_location, contains("bd"))
```
<br><br>

# Grouping Data

Suppose we wanted to know where and when the most amount of battle deaths occured. We could arrange the data to show the most battle death occurrences first. But, if we wanted to know where and when the most battle death occurred (in the aggregate), we should group by region/battle location and year and then tally up the battle deaths to see which region/battle location and year have the most battle death occurrences.  

```{r}
data %>%
  arrange(desc(bd_best)) %>%
  select(conflict_id, battle_location, year, bd_best)

data %>%
  group_by(region) %>%
  tally(bd_best, sort = TRUE)

data %>%
  group_by(battle_location) %>%
  tally(bd_best, sort = TRUE)

data %>%
  group_by(year) %>%
  tally(bd_best, sort = TRUE)

data %>%
  group_by(region, year) %>%
  tally(bd_best, sort = TRUE)
```
<br>

\textit{Why do each of these tables give a different intuition about where and when the most amount of battle deaths occured? It is always important to keep in mind which variables we are using and how we are using those variables.}
<br><br>

# Bar Charts

Suppose we wanted to know the distribution of the types of conflict this dataset contains (i.e. the distribution of the types of conflict that occurred during 1989-2018). We can create a bar chart to visualize this distribution.

```{r}
data %>%
  ggplot(mapping = aes(x = type_of_conflict)) +
  geom_bar()

data %>%
  ggplot(mapping = aes(x = type_of_conflict)) +
  geom_bar() + 
  coord_flip()

data %>%
  mutate(type_of_conflict = case_when(
    type_of_conflict == "2" ~ "interstate",
    type_of_conflict == "3" ~ "internal",
    type_of_conflict == "4" ~ "internationalized internal")) %>%
  ggplot(mapping = aes(x = type_of_conflict)) +
  geom_bar() +
  ggtitle("Distribution of Type of Conflict (1989-2018)") +
  xlab("Type of Conflict") + ylab("Count")
```
<br><br>

# Scatter Plots

Suppose we wanted to see the number of battle deaths across the time period of the dataset (1989-2018) and identify whether there is a pattern. We can create a scatter plot to visualize this. Indeed, with a scatter plot, we can even subset the points by type of conflict, adding another layer of information about the pattern.

```{r}
data %>%
  ggplot(mapping = aes(x = year, y = bd_best, color = type_of_conflict)) + 
  geom_point()
# Factor `type_of_conflict`

data %>%
  ggplot(mapping = aes(x = year, y = bd_best, colour = factor(type_of_conflict))) + 
  geom_point()

data %>%
  ggplot(mapping = aes(x = year, y = bd_best, color = factor(type_of_conflict))) + 
  geom_point() +
  ggtitle("Number of Battle Deaths (1989-2018)") +
  labs(x = "Year", y = "Battle Deaths (best estimate)", color = "Type of Conflict")
```
<br>

Suppose we wanted to know the number of battle deaths across time in Iraq and determine the trend. We can create a scatterplot showing the battle deaths across time for Iraq (even subset the points by type of conflict) and add a trend line.

```{r, message=FALSE, warning=FALSE}
data %>%
  filter(battle_location == "Iraq") %>%
  arrange(conflict_id, year) %>%
  select(conflict_id, battle_location, year, bd_best) %>%
  ggplot(mapping = aes(x = year, y = bd_best)) +
  geom_point(aes(color = factor(conflict_id))) +
  geom_smooth() +
  ggtitle("Battle Deaths in Iraq over the Years (1989-2018)") +
  labs(x = "Year", y = "Battle Deaths in Iraq (best estimate)", colour = "Conflict ID")
```
<br>

*Do it yourself*: How would you see the number of battle deaths across the time period of 1989-2018 and identify whether there is a pattern with respect to the main conflict issue? (hint: `incompatibility`)
<br><br>

# Histograms

Suppose we wanted to visually summarize the number of times battle deaths of minor intensity occurred in any given year (between 25 and 999 battle-related deaths) and see the shape and spread of the freqeuncy of these battle death numbers for each region. We can create histograms that show the number of times battle deaths of minor intensity occurred for each region.

```{r, message=FALSE}
a1 <- data %>%
  filter(bd_best <= 999, region == 1) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +
  ggtitle("Battle Deaths Count in Europe") +
  labs(x = "Battle Deaths (minor)", y = "Count")

a2 <-  data %>%
  filter(bd_best <= 999, region == 2) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +
  ggtitle("Battle Deaths Count in \n the Middle East") +
  labs(x = "Battle Deaths (minor)", y = "Count")
# \n is for new line (R Studio special characters)

a3 <- data %>%
  filter(bd_best <= 999, region == 3) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +
  ggtitle("Battle Deaths Count in Asia") +
  labs(x = "Battle Deaths (minor)", y = "Count")

a4 <- data %>%
  filter(bd_best <= 999, region == 4) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +
  ggtitle("Battle Deaths Count in Asia") +
  labs(x = "Battle Deaths (minor)", y = "Count")

a5 <- data %>%
  filter(bd_best <= 999, region == 5) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +
  ggtitle("Battle Deaths Count in \n the Americas") +
  labs(x = "Battle Deaths (minor)", y = "Count")
# \n is for new line (R Studio special characters)

grid.arrange(a1, a2, 
             a3, a4, 
             a5)

# Comparison between the regions proves difficult because of y axis scaling

b1 <- data %>%
  filter(bd_best <= 999, bd_best >=100, region == 1) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +
  coord_cartesian(ylim=c(0,40)) +
  ggtitle("Battle Deaths Count in Europe") +
  labs(x = "Battle Deaths (minor)", y = "Count")

b2 <-  data %>%
  filter(bd_best <= 999, bd_best >=100, region == 2) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") + 
  coord_cartesian(ylim=c(0,40)) +
  ggtitle("Battle Deaths Count in \n the Middle East") +
  labs(x = "Battle Deaths (minor)", y = "Count")

b3 <- data %>%
  filter(bd_best <= 999, bd_best >=100, region == 3) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +  
  coord_cartesian(ylim=c(0,40)) +
  ggtitle("Battle Deaths Count in Asia") +
  labs(x = "Battle Deaths (minor)", y = "Count")

b4 <- data %>%
  filter(bd_best <= 999, bd_best >=100, region == 4) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +  
  coord_cartesian(ylim=c(0,40)) +
  ggtitle("Battle Deaths Count in Asia") +
  labs(x = "Battle Deaths (minor)", y = "Count")

b5 <- data %>%
  filter(bd_best <= 999, bd_best >=100, region == 5) %>%
  ggplot(mapping = aes(x = bd_best)) +
  geom_histogram(color = "black", fill="white") +  
  coord_cartesian(ylim=c(0,40)) +
  ggtitle("Battle Deaths Count in \n the Americas") +
  labs(x = "Battle Deaths (minor)", y = "Count")

grid.arrange(b1, b2, 
             b3, b4, 
             b5)
```
<br>

*Do it yourself*: How would you visually summarize the number of times battle deaths of major intensity occurred in any given year (greater than 999 battle-related deaths) and see the shape and spread of the freqeuncy of these battle death numbers for each region?
<br><br>

# Box Plots

Suppose we wanted to visually summarize the number of times battle deaths of minor intensity occurred in any given year for each region in a way that would allow for better comparison than the previous histograms. We can create side-by-side boxplots.

```{r}
data %>% 
  filter(bd_best <= 999, region) %>%
  ggplot(mapping = aes(x = factor(region), y = bd_best)) +
  geom_boxplot()

data %>% 
  filter(bd_best <= 999, region) %>%
  ggplot(mapping = aes(x = factor(region), y = bd_best, fill = factor(region))) +
  geom_boxplot() +
  labs(x = "Region", y = "Battle Deaths (Count)") +
  scale_fill_discrete(name = "Region", labels = c("Europe", 
                                                  "Middle East", 
                                                  "Asia", 
                                                  "Africa", 
                                                  "Americas"))
```
<br><br>

# Next Steps

In order to do more advanced modeling and analysis of the data (that would allow for hypothesis testing), one should go deeper into understanding and fixing problems with the dataset (e.g. missing values or NA's), tidying the data, and learnng the R tools that would be best suited for that data (be it `lm()`, `glm()`, `nnet()`, etc.).